\section{Introduction}

Type systems are an essential high level abstraction of modern
programming languages and type checkers are standard components of
their compilers. A language's type system needs to be fitted with a
formal description to allow the language designer to reason about
certain formal aspects of the system, e.g., well-formedness, progress,
and preservation.

Questions about the relationship between a type system's formal
description and its implementation have been addressed in existing
work, ranging from mutual consistency between description and
implementation~\cite{Levin2003,Dijkstra2006} to the development of
specification techniques suitable for generating type
checkers~\cite{Gast04}. Yet the question remains if the standard,
text-book way to formalize type systems -- inference rules over type
judgements -- can be used to derive or generate type checkers.

In this paper we focus on this question and show that a well known
extension of the standard way to formalize type systems, namely the
extension with constraints, can be used to derive type checking
functionality. We develop an abstract type checking framework
utilizing the constraint-based flavor of the type system and present
the design and implementation of a \textsc{Haskell} library which
employs our developed ideas and algorithms for deriving type checkers.

The remainder of this report proceeds as follows. In section 2, we
give a precise overview of constraint-based type systems. Sections 3
and 4 present our two algorithms for generating and solving
constraints. Section 5 describes the design and implementation of the
library. Section 6 and 7 employ our framework to define type checkers
for the languages \textsc{Mini-ML} and \textsc{FeatherweightJava} to
give the reader a better understanding of the usage of the
library. Section 8 presents a framework for visualizing the internals
of the derived type checkers. Sections 9 and 10 describe related work
and discuss some open questions regarding the implementation.

The implementation of our library, its documentation, and all example
type checkers are available from~\cite{TypeCheckLib}.
